---
title: "Phylogenies - Using Street Tree Data "
output: html_notebook
---

## LEARNING WITH TREE DATA 

This could be adapted for many different cities, as there are more and more open source street tree inventories all the time:

https://www.portlandoregon.gov/parks/article/620066 (This is an excel file)

http://data.phl.opendata.arcgis.com/datasets/957f032f9c874327a1ad800abd887d17_0.csv

https://catalog.data.gov/dataset/providence-tree-inventory/resource/d4090ae5-42c1-4352-bd93-f2f64e0a4b2e

Washington D.C.

https://opendata.arcgis.com/datasets/f6c3c04113944f23a7993f2e603abaf2_23.csv

In this Notebook, we'll go back to our initial huge data set of NYC Street trees, and continue learning about trait data using the BIEN database. 

```{r}
#fn <- read.csv("https://data.cityofnewyork.us/api/views/5rq2-4hqu/rows.csv?accessType=DOWNLOAD", header=TRUE)
#write.csv(fn,'2015StreetTreesCensus_TREES.csv')
#fn stands for "file name" which is perhaps a bit too simplified. Let's change that to ALLtrees
ALLtrees <- read.csv('2015StreetTreesCensus_TREES.csv')
#Now let's use a command to examine the dimensions of the data
dim(ALLtrees)
# Wow. There are 683,788 rows in this data set. One row for each street tree. There are also 42 columns.
#What class is this dataset?
class(ALLtrees)
#Let's look through all 42 column headings
colnames(ALLtrees)
# Let's look at the first six lines of data
head(ALLtrees)
```

# GETTING A SPECIES LIST, CHECKING TAXONOMY, PHYLOGENIZING

# This section is very similar to code starting at line 474 in "Inventories by Species" Notebook...
```{r}
library(plyr)
count(ALLtrees$spc_latin)
count(ALLtrees$spc_common)
```

There are 31,619 instances where the species name is missing? Why?

One reason is that dead trees were inventoried (as stumps or dead) but not purposefully left unidentified to species. 

So, we can subset the data to just those species that are alive. While we are at it, we can also limit ourselves to just the first 28 columns of data (the second 28 are a lot about stewardship and management of the trees).
```{r}
dim(ALLtrees)
Strees <- subset(ALLtrees, status=="Alive", select=c(created_at:spc_common, address:y_sp))
Strees <- as.data.frame(Strees)
dim(Strees)
```

Another approach would be just to get rid of all of the instances where there is no species name listed, as follows:

```{r}
dim(ALLtrees)
Strees <- subset(ALLtrees, spc_common!="", select=c(created_at:spc_common, address:y_sp))
Strees <- as.data.frame(Strees)
dim(Strees)
```

This second procedure showed that there were exactly four instances city-wide where there is a missing species name even though the tree is alive. This small number of missing data could cause problems for later analyses, so it's good to have eliminated them from the data set. 

Now let's look at what's left. What species are present? Each species is a level, so let's use the command "level" to look at that. 

```{r}
head(Strees)
levels(Strees$spc_latin);levels(Strees$spc_common)
```

What if we wanted, instead, to get not just a list of each species but how many trees there are of that species as well? This is a frequency table, obtained using the command "count" on the variable that we want to examine. Here, I've done it on the latin name. Would you get the same result if you instead did it on the common name? 

```{r}
freqtable <- count(Strees$spc_latin)
colnames(freqtable)[colnames(freqtable)=="x"] <- "scientificName"
freqtable
```

Now let's look at this list of frequencies:

```{r}
#This would add up all the frequencies of each tree; it should sum to the total number of trees in the city
sum(freqtable$freq)
#How common is the tree that is most common?
max(freqtable$freq)
#How rare is the one that's most rare? 
min(freqtable$freq)
median(freqtable$freq)
#What if we made a histogram?
hist(freqtable$freq)
hist(freqtable$freq, freq=F)
```

## What if we want to look at the list broken down by higher-rank taxa (such as genus or family)?
## We can do this with additional R packages, such as 'taxize', but I also learned about 'traitdataform' which might be useful for working with both trait data and phylogenetic data at the same time. 

```{r}
#library(taxize)
#Another lesson teaches how to do classification functions, from the TAXIZE package.
################ Which lesson teaches them 'taxize'? ################################ Any interest in 'taxastand'?
#This is a new package that looks at trait data and phylogenetic data integrated together.
#install.packages("traitdataform")
library(traitdataform)
```

```{r}
head(freqtable)
class(freqtable)
trees.NYC <- standardize_taxa(freqtable, return = c("phylum", "class", "order", "family", "genus"), resolve_synonyms = TRUE)
head(trees.NYC)
tail(trees.NYC)
```

```{r}
freqtreeclass <- count(trees.NYC$class)
freqtreeclass
freqtreefamily <- count(trees.NYC$family)
freqtreefamily
freqtreegenus <- count(trees.NYC$genus)
freqtreegenus
```

In the past, I've taught Phylomatic, but now there are more recent options, most recently one that includes not just angiosperms and gymnosperms (G.Phylomaker, it uses the zanne2014 tree) but also many ferns and other seedless vascular plants. 

It sounds like the most comprehensive as of mid-2019, and the R package is straightforward. 

Here is a link to the paper http://doi.org/10.1111/ecog.04434

To install a package offered by the developer personally, from their GitHub repository, try this:

```{r}
devtools::install_github("jinyizju/V.PhyloMaker")
```

Or another option is this:

```{r}
install.packages("remotes")
remotes::install_github("jinyizju/V.PhyloMaker")
```

Then load the package:

```{r}
library("V.PhyloMaker")
```

Here is an example from the package developer (i.e., tutorial).

```{r}
getwd()
#example <- read.csv("C:/Users/hcallaha/Documents/TEACHING/19-20/Lab in Plant Diversity/R Notebooks/V.phylomaker/Example_species_list.csv")
tree.a <- phylo.maker(sp.list = example,  tree = GBOTB.extended,  nodes = nodes.info.1, scenarios="S3")
write.tree(tree.a$scenario.3, "Figure.1a.tre")
plot(tree.a$scenario.3, type = "fan")
```

So, above, we generated a list with higher level taxonomy for NYC street trees. Narrowing it to include just species name, genus and family is what this package requires, and I believe the columns where the names of the variables are kept must be in the correct order and named to match the package requirements. 

```{r}
head(trees.NYC)
phylotrees.NYC <- trees.NYC[c(2, 10, 9)]
names(phylotrees.NYC)[1] <- "species"
head(phylotrees.NYC)
```

```{r}
tree.NYC <- phylo.maker(sp.list = phylotrees.NYC,  tree = GBOTB.extended,  nodes = nodes.info.1, scenarios="S3")
#So what this generates is  not a phylo object and not a dataframe, but a two-object list)
#The two objects in this class are, first, the phylo object
# ..... tree.NYC$scenario.3
# The second is a dataframe of the species
# ........... tree.NYC$species.list
```

```{r}
write.tree(tree.NYC$scenario.3, "NYCtree.tre")
#If you look at it in Notepad or similar, you will see that it has the Newick code, including branch lengths throughout the tree.
#We can roughly plot this here, but really will want to use something like FigTree or iTOL to create better figures, so we need to export the file with the tree information.
plot(tree.NYC$scenario.3,no.margin=TRUE,edge.width=2, type = "fan")
```

Now, choose a NYC borough and analyze the NYC Street Tree data to create a phylogeny of all the trees in that single borough.

Hint: go back to the 'Strees' data frame we created earlier, filter it to only represent the trees in your chosen borough, and use the V.PhyloMaker package to create your phylogeny.

Ideally do an analysis PICANTE style, where there is coding of trees by most common.

# EDITS

# Line 40
# Line 106
# Line 193: do you want to make an additional assignment using 'picante'?